---
title: "[Tuto{{< fa brands r-project >}}ial - Part 1]{.no-print} [TutoRial - Part 1]{.print-only}"
subtitle: "Marine Ecosystem Dynamics - 2025"
author: "Kinlan M.G. Jan"
format:
  html:
    html-math-method: katex
    toc: true
    toc-title: "Tuto{{< fa brands r-project >}}ial 1"
    toc-depth: 3
  pdf:
    pdf-engine: weasyprint
css: tutorial.css
format-links: [pdf]
---

::: no-print
## New script

As seen during the presentation, we will keep track of our progress. We thus need to open a new script following one of the option below.

```{mermaid}
flowchart LR
  A[1. File] --> B[2. New File]
  B --> C[3. R script]
```

```{mermaid}
flowchart LR
  A["⌘/Ctlr + ⇧ + N"]
```
:::

## [{{< fa brands r-project >}} syntax]{.no-print} [**R** syntax]{.print-only}

[{{< fa brands r-project >}} is a programming language that use a simplified syntax. In this section, we will explore how to write a script and execute it.]{.no-print} [**R** is a programming language that use a simplified syntax. In this section, we will explore how to write a script and execute it.]{.print-only}

But first some syntax information:

-   Everything after `#` is considered as a comment and will not be executed. It is very important to write what we are doing, so we do not get lost next time we open our scripts.

```{r}
#| echo: true
#| eval: true

# 2 + 2 will not work because of the #
2 + 2 # We should then annotate our script like this
```

-   Several lines of code can be written in one line but must be separated by a semicolon

```{r}
#| echo: true
#| eval: true
2 + 2
3 * 2

# This can also be written as follow:
2 + 2 ; 3 * 2
```

-   [In {{< fa brands r-project >}} we can name any object using `=`, `<-`, `->` or `assign`]{.no-print} [In **R** we can name any object using `=`, `<-`, `->` or `assign`]{.print-only}

```{r}
#| echo: true
#| eval: true
c(1, 2, 3, 4) -> my_first_vector
my_vector <- c(1, 2, 3, 4)
my_function = function(x){x + 2}
assign("x", c(2, 3, 4, 5))
```

-   `==` is a logical function that can be translated as *is equal to*, contrarily *is not equal to* is written `!=`

```{r}
#| echo: true
#| eval: true
2 + 2 == 4
3 * 2 == 4
3 * 2 != 4
```

### Exercises

[Using a new {{< fa brands r-project >}} script, do these calculations:]{.no-print} [Using a new **R** script, do these calculations:]{.print-only}

-   $\ 2^7$

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
2^7
```
:::
:::

-   $\ cos(\pi)$

::: no-print
::: {.callout-tip collapse="true" title="Hint"}
```{r}
#| echo: true
#| eval: false
?cos()
?pi()
```
:::
:::

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
cos(pi)
```
:::
:::

-   The sum of all number from 1 to 100

::: no-print
::: {.callout-tip collapse="true" title="Hint"}
Operations can take place for an entire vector
:::

::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
vector <- seq(from = 1, to = 100, by = 1) # Create a vector from 1 to 100
sum(vector) # Calculate the sum
```
:::
:::

Create a parameter `x1` that equals to 5 and a parameter `x2` that equals to 10

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
x1 <- 5 ; x2 <-  10
```
:::
:::

-   Is $\ 2*$ `x1` equal to `x2`?

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
2 * x1 == x2
```
:::
:::

## Functions

[As seen during the lecture, {{< fa brands r-project >}} works with functions that can:]{.no-print} [As seen during the lecture, **R** works with functions that can:]{.print-only}

-   [Already be implemented in base {{< fa brands r-project >}}]{.no-print} [Already be implemented in base **R**]{.print-only}
-   Comming from another package
-   Created by the user

We will see these three examples in this section, but first it is important to remember that the typical structure of a function is `function(argument1, ...)`.

[Fortunately {{< fa brands r-project >}} helps us to remember what are the needed `arguments`:]{.no-print} [Fortunately **R** helps us to remember what are the needed `arguments`:]{.print-only}

-   Using `help()` or `?`

::: no-print
```{r}
#| echo: true
#| eval: false
help(topic = "sin")
?sin
```
:::

-   Using `example`

::: no-print
```{r}
#| echo: true
#| eval: true
example(sum)
```
:::

For the functions that comes from external packages, we first need to install the new packages. The most common way to do so is by executing `install.packages("Package_Name")`. Then when we want to load the functions, we start the script by executing `library(Package_Name)`.

Finally, if we really do not find a suitable function in a package, we can create your functions following this general structure, but this will not be covered in this tutorial:

```{r}
#| echo: true
#| eval: false
my_function <- function(<argument1>, <argument2>, ...){
  <here comes the definition of my function>
  return(<output of the definition>)
}
```

### Exercises

-   [What is the function `log()` doing and from were does this function come from (base {{< fa brands r-project >}}, other packages)?]{.no-print} [What is the function `log()` doing and from were does this function come from (base **R**, other packages)?]{.print-only}

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: false
?log() #It takes the natural logarithm of the value, it comes from base R
log(10) 
```
:::
:::

-   What are the mandatory arguments for the function `plot()`

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: false
?plot() # the coordinates points x and y are needed
```
:::
:::

-   Is there help associated with the functions from a loaded package?

::: no-print
::: {.callout-tip collapse="true" title="Hint"}
The function `ggplot()` comes from the package `ggplot2`
:::

::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: false
library(ggplot2)
?ggplot # Yes, there is also help for the imported functions
```
:::

::: {.callout title="Optional exercises"}
-   Create a function that print `Hello World!` when executing it

::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
Hello <- function(){
  cat("Hello World!") # print("Hello World!") works too
}
Hello()
```
:::

-   Create a function that multiply the input by 4

::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
multiplyeR <- function(x, y = 4){
  return(x * y)
}
multiplyeR(x = 2) # It works with values
multiplyeR(x = seq(1, 3, 1)) # But also vectors
```
:::
:::
:::

::: print-only
\newpage
:::

## Vectors

[{{< fa brands r-project >}} works with vector from which we can do our calculations.]{.no-print} [**R** works with vector from which we can do our calculations.]{.print-only} Several ways exist to create a vector:

-   Using `c()`, values are added next to each other and separated with a `,`.

```{r}
#| echo: true
#| eval: false
c(1, 2, 1, 4) # It works with integers (round numbers)
c(1.1, 2.4, 3.14652) # It works with floats (decimal numbers)
c("chocolate", "ice-cream") # It works with character
c(TRUE, FALSE) # It works with logical variables
```

-   Using `rep()` to repeat the same values several times.

```{r}
#| echo: true
#| eval: false
rep(x = 3, 2) # it reads: repeat 2 times the value x that is equal to 3
rep(x = "chocolate", 3) # it reads: repeat 3 times the value x that is equal to "chocolate"
```

-   Using `seq()` to create a sequence of values. It only works for numeric values!

```{r}
#| echo: true
#| eval: false
seq(from = 0, to = 10, by = 2) # it reads: create a sequence of values from 0 to 10 every 2 numbers
seq(from = -1, to = 1, by = 0.2) # it also works with negative values and decimal
```

-   Combining all of the above

```{r}
#| echo: true
#| eval: false
rep(x = c(seq(from = 2, to = 3, by = 0.2), 5), 2)
c(rep(x = "character", 5), "other character")
c(seq(from = 2, to = 10, by = 2), rep(x = 1000, 2), c(1, 4, 2))
```

### Exercises

-   Create a vector `v1` that contains the values 1, 2, 3, 4, 6

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
v1 <- c(1, 2, 3, 4, 6)
```
:::
:::

-   Create a vector `v2` that contains 10 times the values 1, 2, 3, 4, 6

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
v2 <- rep(v1, 10)
```
:::
:::

-   Create a vector `v3` that repeats `TRUE`, `FALSE` 2 times

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
v3 <- rep(c(TRUE, FALSE), 2)
```
:::
:::

-   Create a vector `v4` that goes from 10 to 2000

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
v4 <- 10:2000
# or 
v4 <- seq(from  = 10, to = 2000, by = 1)
```
:::
:::

-   Create a vector `v5` that contains `v1`, `v2`, `v3` and 2 times `v4`

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
v5 <- c(v1, v2, v3, rep(v4, 2))
```
:::
:::

::: print-only
\newpage
:::

## Dataframe

Most likely, we will work with data stored in dataframes. A dataframe is composed of observations (rows) and variables (columns). We can see a dataframe like multiples vectors put togethers.

For example in the dataframe below (named `df`) is composed of 4 vectors:

1.  `Species` that contains the species names
2.  `Abundance` that contains the abundances of the species
3.  `Location` that contains the location of the species
4.  `Date` that contains the sampling date

```{r}
#| echo: false
#| eval: true
df <- data.frame(Species = c("Acartia", "Pseudocalanus", "Centropages"),
           Abundance = c(34, 12, 17),
           Location = c("Askö", "Askö", "Askö"),
           Date = c("03-09-2024", "04-09-2024", "02-09-2024"))
print(df)
```

We can access the individual columns (i.e., vectors) using `$`

```{r}
#| echo: true
#| eval: true
df$Species
df$Abundance
df$Location
df$Date
```

### Exercises

-   Create a vector `genus` containing the character `"Acartia", "Centropages", "Temora", "Acartia", "Centropages", "Temora"`

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
genus = c("Acartia", "Centropages", "Temora", "Acartia", "Centropages", "Temora") 
# or genus = rep(c("Acartia", "Centropages", "Temora"), 2)
```
:::
:::

-   Create a vector `station` containing the character `"Askö", "Askö", "Askö", "Tjarnö", "Tjarnö", "Tjarnö"`

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
station = c(rep("Askö",3),rep("Tjarnö", 3))
```
:::
:::

-   Create a vector `abundance` containing the values `3, 10.2, 4, 2.3, 4, 9.4`

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
abundance = c(3, 10.2, 4, 2.3, 4, 9.4)
```
:::
:::

-   Combine all the vectors in a dataframe called `df`

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
df <- data.frame("Genus" = genus,
                 "Station" = station,
                 "Abundance" = abundance)
```
:::
:::

-   Create a vector `output` that correspond to the column `Abundance` of the dataframe `df`. Is `output` similar to the vector `abundance`?

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
output <- df$Abundance # or df[[3]]
output == abundance
```
:::
:::

##  [Importing data in {{< fa brands r-project >}}]{.no-print} [Importing data in **R** ]{.print-only}

::: print-only
More often we enter our data in spreadsheets. We then need to import our data in **R** to process them.<br>
To do so, we use the `read.*` function family.
:::

::: no-print
More often we enter our data in spreadsheets. We then need to import our data in {{< fa brands r-project >}} to process them.<br>
To do so, we use the `read.*` function family.
:::

A typical data import protocol looks like this:

1. Set the working directory with its absolute path

```{r}
#| echo: true
#| eval: false
setwd("/Absolute/Path/To/Working/Directory")
```

2. Import your dataset in your environment

```{r}
#| echo: true
#| eval: false
df <- read.csv("./Relative/Path/Dataset.csv")
```

3. Examine the structure of the data to see if the importation worked well

```{r}
#| echo: true
#| eval: false
str(df)
head(df)
tail(df)
```

### Exercises

```{r, echo=FALSE, results='asis'}
#| echo: false
#| eval: true
#| warning: false
#| message: false

if (knitr::is_html_output()) {
  library("downloadthis");library(PlanktonData)
  data(zooplankton)
  zooplankton  |> 
    download_this(
    output_name = "zooplankton_seasonality",
    output_extension = ".csv",
    button_label = " Click here to download the dataset zooplankton_seasonality.csv",
    button_type = "primary",
    has_icon = TRUE,
    csv2 = FALSE,
    icon = "fa fa-save")
} else {
  cat("You can download the dataset on [GitHub](https://github.com/KMGJan/MED2024/tree/main/MATERIAL/tutorial/assets)")
}
```



* Import the dataset in your environment

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
```{r}
#| echo: true
#| eval: true
df <- read.csv("./assets/zooplankton_seasonality.csv")
```
:::
:::


 * How many rows and columns does this dataset contain?
 
::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}

The structure of the dataset shows that there is 7 variables (columns) and 2956 observations (rows)

```{r}
#| echo: true
#| eval: true
str(df)
```
:::
:::

 * What are the headers of the columns?

::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}

Both the stucture and the head show that the headers are: `Month_abb`, `Year`, `Station`, `Coordinates`, `Group`, `Taxa`, `Biomass`

```{r}
#| echo: true
#| eval: true
head(df)
```

:::
:::

 * What is the last row?
 
::: no-print
::: {.callout-note collapse="true" title="Solution" appearance="simple"}
To see the last row, use the `tail` function

```{r}
#| echo: true
#| eval: true
tail(df)
```
:::
:::